**fabryka**

### 2025-07-14 fabryka_pci/article.Rmd
correction of fabryka article and shiny application after first round of review in PCI archaeology

---
title: '**Fabryka**'
subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
output:
  pdf_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Fabryka}
- \fancyhead[R]{R web application for fabric analysis}
- \fancyfoot[C]{\thepage}
- \usepackage{float}
- \floatplacement{figure}{H}
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

@@

Marc THOMAS ^1\*^ (https://orcid.org/0000-0002-8160-1910) \smallbreak

@@

**Key words** : R shiny web application, fabric, site formation
processes, spatial analyses, orientation and inclination \bigbreak

@@

This article presents Fabryka, an open access R shiny web application
dedicated to the fabric analysis of archaeological or geological
assemblages. The fabric is a statistic made of the orientation and
inclination of elongated sedimentary particles and archaeological
objects. It provides information about the source of sediments, the
nature of the geological processes involved in the formation of the
deposits and their impact on the preservation of the archaeological
assemblages. These statistical analyses are not easily accessible to
non-specialists and are particularly time-consuming, sometimes requiring
the use of several software, some of which are chargeable. Fabryka uses
classical analytical procedures (Benn, rose and Schmidt diagrams and
statistical tests on orientations) for analysing the fabrics of objects
and includes a spatial method. New statistical methods and new ways of
exploration are proposed in the application and this present article.
Also, a call for contributions is proposed to construct a new reference
model with spatial information. \smallbreak

You can use Fabryka $\beta$ 3.0 by following this link
<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
updated to Fabryka 1.0 after the review process of this article. The
Fabryka application files, made in R [@rcoreteam2025] by using the R package *shiny* [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak

@@

Fabric is made of the orientation and inclination of elongated
sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It is
a statistic (*i.e.*, it needs as much measurements as possible) that
enables the identification of sediment sources that make up the deposits
[@texier2020], sedimentary processes involved in their formation
[@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be treated as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, the fabric analysis
made of the sedimentary deposits depends mainly on the archaeological
material (bones and lithics), which enables them to be inserted into the
formation dynamics of the site and to obtain information on their
representativeness for the reconstruction of past activities.

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the direction of flow, with a slight upstream
inclination. Since G. Isaac's pioneering work, there has been an
abundant literature devoted to the analysis of fabrics [@benn1994;
@lenoble2004; @mcpherron2018] allowing this method to be increasingly
performed on archaeological sites. Several software allow data
processing, some of which are not freely accessible, do not allow
spatial exploration of the data or do not offer interfaces to facilitate
the task of non-programmer users [@mcpherron2018].

The following sections explain all the selected approaches and our
proposals for processing the fabrics thanks to fabryka $\beta$ 3.0, from
taking measurements in the field, through the statistical methods and
graphical modes of representation used in the application (fig.
\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/presentation_applications.png}
  \caption{Fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

@@

During excavation, the fabric of elongated remains is generally
systematically measured. The direction (= bearing) of the lowest point
is measured in the range [0; 360] with North as the reference (*i.e.*
direction = 0°). The inclination (= plunge, dip) is measured in the
range [0; 90] with horizontal as a reference. The direction [0; 360] can
be converted into an orientation measured in the range [0; 180] by
subtracting 180° from the measurements in the range ]180; 360].

Traditionally, the direction of the objects lowest point is measured
using a compass and the inclination is measured using an inclinometer.
The aim is to measure the direction and inclination of the physical axis
of elongation of objects ignoring its shape. To do this, the measuring
device must be placed parallel to the general axis of elongation of the
piece without touching it. Measuring the fabric from its imprint or
visible surface is biased by its shape, making impossible to assess its
true inclination. Since the theodolite was introduced to excavations in
the 1990s, the fabrics is sometimes measured by recording the
coordinates of the ends of objects. In the latter case, the rules of
trigonometry are applied to calculate directions and inclinations.

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>) which can
be used to measure fabrics [@discamps2024]. The device stores and
associates data with the order of measurement. When the device is tilted
downwards, the direction of the lowest point is measured and, the
recorded inclination is negative. When the device is tilted upwards, the
direction of the highest point is recorded, and the inclination is
positive. To obtain data that is suitable for statistical analysis, in
the first case the absolute value of the inclination must be retained.
In the second case (*i.e.* positive inclination), if the measured
direction is in the interval [0; 180], 180° must be added to the
direction, and if the measured direction is in the interval ]180; 360],
180° must be subtracted from the direction.

When possible, preferring the use of a digital inclinometer avoids many
errors during measurement and data recording. On the other hand, for the
reasons given above, the use of a theodolite to measure fabrics is not
recommended as it results in inaccurate inclination measurements. As
well, measuring orientations and inclinations in sections leads to
sampling biases.

A sample corresponds to all the direction and inclination measurements
taken within a Lithostratigraphic Unit (UL). UL seems to be the most
coherent entity which needs to be explored because they are defined
according to sedimentary criterion, which a priori guarantee a certain
coherence in terms of the formation dynamics we are seeking to approach.
In Fabryka, analyses are constrained within samples unless the user
decides otherwise in the spatial exploration panel. \bigbreak

@@

Two main statistical methods are used:

1.  The first considers the orientation only. The calculation of the
    orientation rate 'r' (or 'L' when expressed as a percentage,
    Curray's [-@curray1956] formula tests the hypothesis of a unimodal
    distribution of orientations. If the 'p' value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the orientation rate 'r' where '$\alpha$' is
    the orientation in radians and 'n' is the sample size is as follows:

Curray's formula to calculate 'L' is (**1.**)

$$L = \frac{r*100}{n}$$ with (**2.**)

$$r = \sqrt{\sum_{i=\alpha}^n sin{(2\alpha)} + \sum_{i=\alpha}^n cos{(2\alpha)}}$$
\smallbreak where 'n' is the number of measures and '$\alpha$', is the
orientation (0 to 180°) \smallbreak

The Rayleigh test is performed thanks to the formula (**3.**)

@@

The method consisting of doubling the value of the orientations before
calculating the orientation rate L and the Rayleigh test, as proposed by
Krumbein [-@krumbein1939], tests the hypothesis of a bimodal distribution
of orientations with a period of 90°. The hypothesis of a 90° period
bimodal distribution of a sample of orientations is retained when p \<
0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
size 'n' prevents samples from being compared with each other when 'n'
varies [@curray1956; @bertran2002]. For example, when the orientation
rate "L" is 25%, the Rayleigh test gives p-values below a critical
threshold of 0.05 for series with more than 50 measurements (fig. 2).

2.  The second considers the direction and inclination of the objects.
    The normalised eigenvalues E1, E2 and E3 are calculated from the
    cosine formed by the axis of the objects with the three axes of
    space [@watson1966]. The Benn indices [@benn1994] describe the
    fabric shape of a sample. The elongation index EL is equal to
    1-(E2/E1) and the isotropy index IS is E3/E1. When EL is high, the
    axes of the objects are grouped around a direction. When IS is high,
    the fabric is said to be isotropic meaning the axes of the objects
    show no order. Finally, when the value 1-EL-IS is high, the fabric
    is said to be planar, with the axes of the objects grouped around a
    plane. \bigskip

```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}

@@

ggplot(rayleigh_table, aes(x=L)) + 
    geom_line(aes(y = p_value_sim_100, colour = "n = 100"), size = 1) +
    geom_line(aes(y = p_value_sim_30, colour = "n =  30"), size= 1) +
    geom_line(aes(y = p_value_sim_40, colour = "n =  40"), size= 1) + 
    geom_line(aes(y = p_value_sim_50, colour = "n =  50"), size= 1) +
    geom_line(aes(y = seuil_0.05, colour = "threshold 0.05"), size= 0.5) +
    xlab("orientation rate 'L' (%)") +
    ylab("p-value") + 
    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60)) +
scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1)) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  labs(colour = "")

@@

To distinguish different statistical populations, a spatial method
proposed by McPherron (2018) is used. Thanks to this method it is
possible to calculate the Benn indices and to test the orientation rate
'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
remains, and its n nearest neighbours in 3 dimensions (McPherron
[-@mcpherron2018] R script, modified , formula (**4.**)). The analysis
of the fabric as proposed by McPherron [-@mcpherron2018] has the
advantage of spatialising the measurements and thus avoiding the
arbitrary sampling classically carried out by square and décapage. It
is possible to localize the statistical signal of fabrics by varying the
number of measurements sampled to calculate the Benn indices. The choice
of about 40 to 60 measurements (each remain and its 40 to 60 nearest
neighbours) to calculate the Benn indices allows local particularities
to be observed, and above all, it allows the Rayleigh test to be carried
out in its field of application (stability of the test between around 40
to 50 measurements after @bertran2002).

By avoiding the instability of the Rayleigh test, the rejection rate
'Tr' of the null hypothesis provides a more reliable description of the
distribution of orientations in a sample. In addition, it allows
comparisons between samples despite varying sample size.

The rejection ratio 'Tr' is performed thanks to the formula (**4.**)

@@

Several methods of representation are used. Rose diagrams summaries the
orientation and inclination measurements for each sample. Benn indices
are projected in a Benn diagram and measures (directions and plunge) are
projected into Schmidt diagrams. These figures are available for both
'classical' and 'spatialised' methods.

As proposed by @mcpherron2018, a spatial investigation of the fabric
is proposed in a dedicated panel. Benn indices (1994) calculated for
each spatial series of remains are projected within a Benn diagram in
which each pole (isotropic, linear, and planar) is coloured respectively
in red, green and blue. The RGB colour code is reused to materialize the
fabric of the remains on spatial projections (plan and section views) of the material along the axes of the excavation grid. \bigbreak

@@

The next sections provide a description and a tutorial of Fabryka.

@@

The simplest method to upload data in the application is to name the column names directly, as shown in the example files. If you miss a context column (*i.e.* 'id', 'code' or 'sample'), please create it by repeating a same character. By doing so, you do not need to select your variables in the main panel. The application correctly reads the data file directly.

@@

Fabryka supports three angle measurement methods:

@@

-   check 'Case 1: Only angles' if you have in your dataset the columns
    'id', 'sample', 'code', direction' and 'dip'.
-   check 'Case 2: Only angles from DistoX2' if you have in your dataset
    the same columns that above but your data comes from a DistoX2 and
    needs to be converted.
-   check 'Case 3: Angles from DistoX2 with coordinates' if you have in
    your dataset the same columns that above (Case 2, data from DistoX2)
    with a single set of coordinates (x, y, z).
-   check 'Case 4: Angles and coordinates' if your data has the same
    form that above (Case 3) but is coming from a compass and a
    clinometer.
-   check 'Case 5: Two shots data without angles' if you have the
    columns 'id', 'sample', 'code' and the angles have been measured
    with a total station (*i.e.* two shots data).

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a dataset coming from
Le Moustier Lower shelter (Dordogne, France). The example files directly
contain the right column names. You can use them directly in the
application. Whatever the names of the columns in your file, you can
select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(*i.e.* no empty modalities, so no line without a fabric measurement).
When a data table with new columns appears below the column selector,
you are ready to use the application.

If your data comes from a DistoX2, the columns 'direction' (= 'bearing',
0 to 360° angles) and 'dip' (= 'plunge') are automatically corrected in
the new table. Whatever the form of your dataset, two new columns
'orientation_pi' (0 to 180° angles) and 'angle_double' (see Krumbein
1939) appear. If the angles were measured using a total station (*i.e.*
two shots data), the bearing and the plunge are also calculated. If only
one set of coordinates is provided with angles, a second set of
coordinates is calculated and displayed in the new table (fig.
\ref{fig:figure_upload_your_data}G). You can download the new data by
clicking the 'copy', the 'csv' or the 'pdf' buttons. \bigbreak

@@

@@

-   Benn diagrams with possible integration of @bertran2002 model (fig. \ref{fig:figure_classical_method}D)
-   Rose diagrams for bearing and plunge (fig. \ref{fig:figure_classical_method}G)
-   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
-   Summary data table with Benn indices, additional statistics, and orientation statistical tests.

@@

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues (see Woodcock 1977). They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axis of the space. They are used to compute
    the Benn indices
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles [@krumbein1939].

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters ('filter by sample'
and 'filter by code') and the slider (*i.e.* minimum number of measures) in the left
sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
Certain geological processes can have a different impact on objects
depending on their physical properties (*e.g.* silex versus bones in a
water flow as in the example files provided where bones have a strong
preferential orientation and are much more 'linear' in the Benn diagram,
see @thomas2019 and @texier2020). The natural processes models that
you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel. In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations [@curray1956] in the
summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).

>>>>

---
title: '**fabryka**'
subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
output:
  pdf_document: default
header-includes:
- \usepackage{amsmath}
- \usepackage{lineno}
- \linenumbers
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{fabryka}
- \fancyhead[R]{R web application for fabric analysis}
- \fancyfoot[C]{\thepage}
- \usepackage{float}
- \floatplacement{figure}{H}
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

\renewcommand{\[}{\begin{equation}}
\renewcommand{\]}{\end{equation}}
```{=html}
<style>
body {
text-align: justify}
</style>
```

@@

Marc THOMAS ^1\*^ (<https://orcid.org/0000-0002-8160-1910>) \smallbreak

@@

**Key words** : R shiny web application, fabric, site formation
processes, spatial analysis, orientation and dip \bigbreak

@@

This article presents fabryka 1.1, an open access R shiny web
application dedicated to the fabric analysis of archaeological or
geological assemblages. The fabric is a statistic made of the
orientation and dip of elongated particles and archaeological objects.
It provides information about the nature of the geological processes
involved in the formation of the deposits and their impact on the
preservation of the archaeological assemblages. These statistical
analysis are not easily accessible to non-specialists and are
particularly time-consuming, sometimes requiring the use of several
software, some of which are chargeable. fabryka uses classical
analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical
tests on orientations) for analysing the fabrics of objects and includes
a spatial method. New statistical methods and new ways of exploration
are proposed in the application and this present article.\smallbreak

You can use fabryka by following this link
<https://marchaeologist.shinyapps.io/fabryka/> or by launching the
application in Rstudio thanks to its script. The fabryka application
files, made in R [@rcoreteam2025] by using the R package *shiny*
[@chang2024], and this article, written in rmarkdown [@allaire2014;
@xie2018; @xie2020], are both available online in a public github
repository (<https://github.com/marchaeologist/fabryka/tree/main>).
\pagebreak

@@

Fabric is made of the orientation and dip of elongated objects
[@bertran1997; @bertran2002; @lenoble2004]. It is a statistic (*i.e.*,
it needs as much measurements as possible) that enables the
identification of sedimentary processes involved in the formation of the
deposits [@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be considered as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, analysis of the
fabric of sedimentary deposits depends mainly on archaeological material
(bones and lithics). This makes it possible to identify the sedimentary
processes to which they were subjected and to discuss their degree of
spatial disorganization.

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the orientation of flow, with a slight upstream dip.
Since G. Isaac's pioneering work, there has been an abundant literature
devoted to the analysis of fabrics [@benn1994; @lenoble2004;
@mcpherron2018] allowing this method to be increasingly performed on
archaeological sites.

Numerical methods have been developed to measure the orientation and dip
of objects in the field [@mcpherron2005artifact; @discamps2024] or from
field archives such as plans and photos [@de2013application;
@sanchez2016assessment].

Two main families of statistical method are commonly used to
characterize the fabric of a given assemblage [@bertran2002;
@mcpherron2018; @de2021new].

The first methods focus on the orientation
of the measurements only and aim to test the null hypothesis of a random
orientation from a uniform distribution through several statistical
tests [@krumbein1939; @curray1956; @rao1967large].

The second statistical method take into account object orientation and
dip. Three normalized eigenvalues (E1, E2 and E3) are calculated from
the cosine formed by the axis of the artifact with the three axes of
space [@watson1966]. @woodcock1977 and @benn1994 indices are used to
describe the fabric shape. Several authors *e.g.*, [@bertran1997;
@bertran2002; @lenoble2004] built models to compare the fabric of
archaeological assemblages to natural sedimentary processes and human
processes (*i.e.*, fabric of knapping spots). The null hypothesis of an
absence of perturbation is a fabric shape compatible with that of
experimental knapping spots. These methods are described in the
Mathematics section of this article.

Several software allow data processing, some of which are not freely
accessible, do not allow spatial exploration of the data or do not offer
interfaces to facilitate the task of non-programmer users
[@mcpherron2018].

The following sections explain all the selected approaches and our
proposals for processing the fabrics thanks to fabryka, from taking
measurements in the field, through the statistical methods and graphical
modes of representation used in the application (fig.
\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/presentation_applications.png}
  \caption{fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

@@

During excavation, the orientation and dip of elongated remains (*i.e.* a > 2 * b and with a > 2 cm) is
measured on a representative set of artefacts or on the whole
assemblage. The orientation (= bearing) of the lowest point is measured
in the range [0; 360] with North as the reference (*i.e.* orientation =
0°). The dip (= plunge) is measured in the range [0; 90] with horizontal
as a reference. The orientation [0; 360] can be converted into axial
data measured in the range [0; 180] by subtracting 180° from the
measurements in the range ]180; 360].

Traditionally, the orientation of the objects lowest point is measured
using a compass and the dip is measured using an inclinometer. The aim
is to measure the orientation and dip of the physical axis of elongation
of objects ignoring its shape. To do this, the measuring device must be
placed parallel to the general axis of elongation of the object.
Measuring the fabric from its imprint or visible surface is biased by
its shape, making impossible to assess its true dip. Since the
theodolite was introduced to excavations in the 1990s, the fabrics is
sometimes measured by recording the coordinates of the ends of objects [@mcpherron2005artifact].
In the latter case, the rules of trigonometry are applied to convert the
pair of coordinates into orientations and dips.

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>) which can
be used to measure fabrics [@discamps2024]. The device stores and
associates data with the order of measurement. When the device is tilted
downwards, the orientation of the lowest point is measured and, the
recorded dip is negative. When the device is tilted upwards, the
orientation of the highest point is recorded, and the dip is positive.
To obtain data that is suitable for statistical analysis, in the first
case the absolute value of the dip must be retained. In the second case
(*i.e.* positive dip), if the measured orientation is in the interval
[0; 180], 180° must be added to the orientation, and if the measured
orientation is in the interval ]180; 360], 180° must be subtracted from
the orientation.

When possible, preferring the use of a digital inclinometer avoids many
errors during measurement and data recording. On the other hand, for the
reasons given above, the use of a theodolite to measure fabrics is not
recommended as it results in inaccurate dip measurements. As well,
measuring orientations and dip in sections leads to sampling biases.

A sample corresponds to all the orientation and dip measurements taken
within a Lithostratigraphic Unit (LU) or a pedostratigraphic unit. These
units seem to be the most coherent entity which needs to be explored
because they are defined according to geological criterion, which a
priori guarantee a certain coherence in terms of the formation dynamics
we are seeking to approach. In fabryka, analysis are constrained within
samples unless the user decides otherwise in the spatial exploration
panel. \bigbreak

@@

Two main statistical family methods are used:

1.  The first considers the orientation only. The calculation of the
    Vector Magnitude $L$ (expressed as a percentage) by Curray's
    [-@curray1956] formula tests the hypothesis of a unimodal
    distribution of orientations. If the $p$ value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the Vector Magnitude $L$ is as follows:

Curray's formula to calculate $L$ is:

$$L = \frac{r*100}{n}$$ with:

$$r = \sqrt{\sum_{i=\alpha}^n \sin{(2\alpha)} + \sum_{i=\alpha}^n \cos{(2\alpha)}}$$
\smallbreak where $n$ is the number of measures and $\alpha$, is the
axial data (0 to 180°) \smallbreak

The Rayleigh test is performed thanks to the formula:

@@

The method consisting of doubling the value of the axial angles before
calculating the Vector Magnitude $L$ and the Rayleigh test, as proposed by
Krumbein [-@krumbein1939], tests the hypothesis of a bimodal
distribution with a period of 90°. The hypothesis of a 90° period
bimodal distribution of a sample of orientations is retained when $p$
\<0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
size $n$ prevents samples from being compared with each other when $n$
varies [@curray1956; @bertran2002]. For example, when the Vector
Magnitude $L$ is 25%, the Rayleigh test gives $p-values$ below a critical
threshold of 0.05 for series with more than 50 measurements (fig. 2).

The Rao [-@rao1967large] spacing test assesses the uniformity of
circular data. It determines whether the data show significant
directionality (*i.e.* the null hypothesis is a uniform distribution).
It is widely used to detect multimodal directionality.

2.  The second considers the orientation and dip of the objects. The
    normalised eigenvalues $E1$, $E2$ and $E3$ are calculated from the
    cosine formed by the axis of the objects with the three axes of
    space [@watson1966].

The fabric shape can be described by calculating the ratios $r1$ = ln(E1/E2), $r2$ = ln(E2/E3) and $K$ = r1/r2 [@woodcock1977]. When 0 \< $K$ \< 1, the fabric is planar. When 1 \< $K$ \< $+\infty$, the fabric is linear. The parameter $C$ = ln(E1/E3) is a measure of the strength of the preferred orientation [@woodcock1977]. These indices can be plotted into a Woodcock diagram.

The Benn indices [@benn1994] describe the fabric shape of a sample and can be plotted in a ternary diagram (*i.e.* Benn diagram). The elongation index EL is equal to $1-(E2/E1)$ and the isotropy index $IS$ is $E3/E1$. When $EL$ is high, the axes of the objects are grouped around an orientation. When $IS$ is high, the fabric is said to be isotropic meaning the axes of the objects are uniformly distributed in all directions. Finally, when the value $1-EL-IS$ is high, the fabric is said to be planar, with the axes of the objects grouped around a plane. \bigskip

```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the Vector Magnitude 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}

@@

ggplot(rayleigh_table, aes(x=L)) + 
    geom_line(aes(y = p_value_sim_100, colour = "n = 100"), size = 1) +
    geom_line(aes(y = p_value_sim_30, colour = "n =  30"), size= 1) +
    geom_line(aes(y = p_value_sim_40, colour = "n =  40"), size= 1) + 
    geom_line(aes(y = p_value_sim_50, colour = "n =  50"), size= 1) +
    geom_line(aes(y = seuil_0.05, colour = "threshold 0.05"), size= 0.5) +
    xlab("orientation rate 'L' (%)") +
    ylab("p-value") + 
    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         
    scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60), expand = c(0, 0)) +
    scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1), expand = c(0, 0)) +
    theme_minimal() +
    theme(legend.position = "bottom") + 
    labs(colour = "")

@@

To distinguish different statistical populations, a spatial method
proposed by McPherron [-@mcpherron2018] is used. Thanks to this method
it is possible to calculate the Benn indices and the Vector Magnitude
$L$ (formula (**3.**) modified from McPherron [-@mcpherron2018] R
script) for each remains, and its $n$ nearest neighbors in 3 dimensions
(McPherron [-@mcpherron2018] R script, modified, formula (**4.**)). The
analysis of the fabric as proposed by McPherron [-@mcpherron2018] has
the advantage of spatialising the measurements. It is possible to
localize the statistical signal of fabrics by varying the number of
measurements sampled to calculate the Benn indices. The choice of about
40 to 60 measurements (each remain and its 40 to 60 nearest neighbors)
to calculate the Benn indices allows local particularities to be
observed, and above all, it allows the Rayleigh test to be carried out
in its field of application (stability of the test between around 40 to
50 measurements after @bertran2002).

By avoiding the instability of the Rayleigh test, the rejection rate
$Tr$ of the null hypothesis provides a more reliable description of the
distribution of orientations in a sample. In addition, it allows
comparisons between samples despite varying sample size.

The rejection ratio $Tr$ is performed thanks to the formula:

@@

Several methods of representation are used. Rose diagrams summaries the
orientation and dip measurements for each sample. Benn indices are
projected in a Benn diagram and measures (orientation and dip) are
projected into Schmidt diagrams. These figures are available for both
'classical' and 'spatialised' methods.

As submited by @mcpherron2018, a spatial investigation of the fabric is
proposed in a dedicated panel. @benn1994 indices calculated for each
spatial series of remains are projected within a Benn diagram in which
each pole (isotropic, linear, and planar) is colored respectively in
red, green and blue. The RGB color code is reused to materialize the
fabric of the remains on spatial projections (plan and section views) of
the material along the axes of the excavation grid. \bigbreak

@@

The next sections provide a description and a tutorial of fabryka. All
the figures in this section were produced using the example data set
provided in the application (*i.e.* "Case 3: Angles from DistoX2 with
coordinates).”

@@

The simplest method to upload data in the application is to name the
column names directly, as shown in the example files. Enter your data in
a csv file, adding the columns 'id', ‘code’ and 'sample'. If the columns
don't exist, fill them in with the same letter. By doing so, you do not
need to select your variables in the main panel. The application
correctly reads the data file directly.

@@

fabryka supports three angle measurement methods:

@@

-   check 'Case 1: Only angles' if you have in your data set the columns
    'id', 'sample', 'code', 'orientation' and 'dip'.
-   check 'Case 2: Only angles from DistoX2' if you have in your data
    set the same columns that above but your data comes from a DistoX2
    and needs to be converted.
-   check 'Case 3: Angles from DistoX2 with coordinates' if you have in
    your data set the same columns that above (Case 2, data from
    DistoX2) with a single set of coordinates (x, y, z).
-   check 'Case 4: Angles and coordinates' if your data has the same
    form that above (Case 3) but is coming from a compass and an
    inclinometer.
-   check 'Case 5: Two shots data without angles' if you have the
    columns 'id', 'sample', 'code' and the angles have been measured
    with a total station (*i.e.* two shots data).

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a data set coming from
the Lower shelter of Le Moustier [@thomas2019; @texier2020]. The example files
directly contain the right column names. You can use them directly in
the application. Whatever the names of the columns in your file, you can
select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(*i.e.* no empty modalities, so no line without a fabric measurement).
When a data table with new columns appears below the column selector,
you are ready to use the application.

If your data comes from a DistoX2, the columns 'orientation' (=
'bearing', *i.e.* 0 to 360°) and 'dip' (= 'plunge') are automatically
corrected in the new table. Whatever the form of your data set, two new
columns 'orientation_pi' (0 to 180° angles) and 'angle_double' (see
Krumbein 1939) appear. If the angles were measured using a total station
(*i.e.* two shots data or 2 by 2 coordinates of plotted pieces refits), the orientation and the dip are calculated. If only one set of coordinates is provided with angles, a second set of
coordinates is calculated and displayed in the new table (fig.
\ref{fig:figure_upload_your_data}G). You can download the new data by
clicking the 'copy', the 'csv' or the 'pdf' buttons. \bigbreak

@@

\begin{figure} [H]
\centering
  \includegraphics[width=16cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

@@

-   Benn diagrams with possible integration of @bertran2002 model (fig.
    \ref{fig:figure_classical_method}D)
-   Rose diagrams for axial data and dip (fig.
    \ref{fig:figure_classical_method}G)
-   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
-   Woodcock diagrams
-   Summary data table with Benn indices, additional statistics, and
    orientation statistical tests.

@@

-   $N$ stands for number of measures in the samples
-   $E1$, $E2$ and $E3$ are the eigenvalues [@watson1966]. They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axis of the space. They are used to compute
    the Benn indices
-   $IS$ and $EL$ are the Benn indices ratios standing
    respectively for, 'isotropy' with $IS = E3/E1$ and 'elongation' with
    $EL = 1-(E2/E1)$ [@benn1994].
-   $K$ and $C$ are the @woodcock1977 indices.
-   $L$ is the Vector Magnitude (strength of the preferred orientation)
    and $R.p$ is the p-value result of the Rayleigh test.
-   $L.double$ and $R.p.double$ are the same statistics that above for
    double angles [@krumbein1939].
-   $Rao_p$ is the p_value of the @rao1967large test.

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters ('filter
by sample' and 'filter by code') and the slider (*i.e.* minimum number
of measures) in the left sidebar panel (fig.
\ref{fig:figure_classical_method}A, B and C).

\begin{figure} [H]
\centering
  \includegraphics[width=16cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analysis, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. Rose diagrams subpanel, H. Schmidt diagram subpanel, I. Summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}
